---
title: "Discrimination simulation"
author: "Chris Berg"
date: "1/15/2020"
output: html_document 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(cache = TRUE)
```

# Introduction 

The purpose of this simulation (inspired by a similar simulation done in Stata by Dr. Scott Cunningham of Baylor Univ.) is to illustrate how some controls can seem sensible to include in a regression, but actually be produce completely "wrong" results (or, specifically, produce estimates biased in a direction *opposite* of the causal relationship).


# Data Generating Process

Let's consider a situation in which wages depend on **unobserved** ability $A_i$ and some measure of status $C_i$. Ability further affects this standing as well, but so does discrimination $D_i$. For individuals facing discrimination ($D_i = 1$), standing is otherwise lower than it would be if they were a "non-oppressed"-type. an important point is, regardless of how one interprets the model, ability affects both status/standing as well as wages on the job-- no matter who you are, if you have higher ability you can get a better job *and* get paid more.

In practice, this could perhaps be the labor market for science or engineering; fields where women have faced sexism and unwelcomeness from the academy. In this case $C_i$ would be some measure of "education status," and women would be less inclined ($\bar C_D < \bar C$ where $\bar C_D$ is the group average education for the discriminated group) to achieve educational standing in these fields. Alternatively $C_i$ could represent some measure of "occupational standing" where under-represented groups are more-inclined to be "shut-out" from higher-status jobs, even if they have the same innate ability as the rest of the population.

The model is thus:

$$ C_i = \mu_c + \beta_D D_i + \beta_A A_i + \epsilon_i$$
$$ w_i = \mu_w + \alpha_C C_i + \alpha_A A_i + \nu_i$$
[**Note, we are ruling-out a channel from wages back to status/standing.** Including such a link would of course produce a reduced-form dependency of ($w$ , $C$) on ($D$ , $A$).]


# Simulation model

### Parameter setup
```{r parameters}
a_c = 0 # "status" intercept
a_w = 10 # wage intercept
sd_c = 1 # "randomness" alt. role of luck in status
b_disc = 1 # parameter governing discrimination penalty (0-1)
b_c_disc = -1 # discrimination hurts your status
b_c_ability = 2 # natural ability determines status
b_w_c = 2 # status leads ot higher wages
b_w_ability = 2 # talent pays off 

D = rbinom(10000 , 1 , 0.5) # half the population is subject to discrimination in this example
ability = rnorm(10000)
discrimination = b_disc*D # targets of discrimination
```

The following is the code for the equations shown in the DGP section above:

```{r models}
C = a_c + b_c_disc*discrimination + b_c_ability*ability + rnorm(10000, mean = 0 , sd = sd_c) # top eqn above
wage = a_w + b_w_c*C + b_w_ability*ability + rnorm(10000)  # bottom eqn above
```

# Analysis

First, just to see that there's "no funny business," let's look at the results from our "God regression" that includes unobserved ability-- which we will of course rule-out the econometrician being able to estimate.

```{r godreg}
true_wage = lm(wage ~ C + ability ) # God regression

summary(true_wage)

```

As we can clearly see, it correctly estimates the DGP for wages.

What about an econometrician tasked with measuring discriminatory wage gaps? Their first instinct might be to look at the raw wage gap between the discriminated group and the rest of the population by regressing wages on an indicator for being in the discriminated class:

```{r ecreg }
ecms_wage = lm(wage ~ discrimination ) # non-Godly regression

summary(ecms_wage)
```

This correctly estimates what you'd if you plugged-in the status variable into wage, and estimated the effect of discrimination. This is what the media might quote as the "discrimination gap"-- the mean difference in earnings between the discriminated and non-discriminated class.

What if someone told the econometrician "Actually, members of the discriminated class simply *appear* to have a lower wage, but they just *choose* to [not get a high-status education; go into a high-status occupation; etc.] so you have to control for $C_i$!" Suppose, then, the econometrician estimates the wage equation with $C_i$ included.

```{r badreg}
ecms_wage_bad = lm(wage ~ discrimination + C) 

summary(ecms_wage_bad)
```

We've estimated a *positive* effect of being in a discriminated group! But this clearly isn't right because as the Gods of this simulation we **know** that this is false.

This illustrates what happens when one *conditions on a "collider."* Here, $C_i$ (education; occupation; etc) is called a "collider" (see Scott Cunningham's *Mixtape* for a deeper treatment). The intuitive reason behind this bias is that we could've conditioned on status *or* a dummy for discriminated group membership, but when we include both, we force our model to compare discriminated and non-discriminated individuals with the same fixed level of status $\tilde C$. Yet, we know that the only individuals who face discrimination but could get the same $\tilde C$ as the non-discriminated, are those who have a much higher average ability than the non-discriminated group... but these are the very people who also receive a higher wage! In this light, we can recognize "it"collider conditioning" as a self-inflicted form of selection bias.

